---
title: "Project 1"
author: "Heather Copley & Andy Johnson"
date: "2023-08-29"
output: 
    html_document:
        toc: true
        toc_float:
            toc_collapsed: true
        code_folding: "show"
        theme: flatly
        highlight: haddock
        df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)

```


# Objective 

The goal of this project is to write functions that will manipulate and process data sets that come in a certain form. 

## Data Processing without Functions

First we will parse and transform the data without creating functions to better understand the tasks we want to perform.
One of the data sets can be read from csv from the following link:

```{r read_in_data}

sheet1 <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv") 

```

### Select Columns
  
We want to select the following columns:  

* Area_name (rename as area_name)  
* STCOU  
* Any column that ends in "D"  

```{r selections}

#select area_name, STCOU, and columns ending with D
data_selections <- sheet1 %>%
    select(area_name = Area_name, STCOU, ends_with('D')) 

```

### Pivot Data from Wide to Long 

Next we pivot the data from a wide format to a long format where each row has only one enrollment value for the area.

```{r pivot_data}

#pivot columns ending with D into 2 columns 
#one for the enrollment code and 
#one for the values
data_pivot <- data_selections %>%
    pivot_longer(cols = ends_with('D'), 
                 names_to = 'Enrollment_code', 
                 values_to = 'Enrollment_value') 

```

### Extract Year and Measurement

Now we extract the year and the measurement from the enrollement code.  

The enrollment code follows a pattern:  

* The first three characters are the survey
* The next four are the type of value from the survey
* The last two digits prior to the "D" represent the year

```{r extract_measurement_year}

#extract the year (digits 8 & 9) and format as a 4 digit year
#extract the measurement (the first 7 characters)
data_extract <- data_pivot %>%
    mutate(year = year(as.Date(substr(Enrollment_code, 8,9), format = '%y')),
           measurement = substr(Enrollment_code, 1, 7))

```

### Split Data

Next we split the data into two data sets, one for county and one for non-county data.
We also assign classes to each data set.


```{r split_data}
#get the county data 
county_data <- data_extract %>%
    filter(str_detect(area_name, ',' ))
#assign county class
class(county_data) <- c('county', class(county_data))

#get the non-county data
non_county_data <- data_extract %>%
    filter(!str_detect(area_name, ',' )) 
##assign state class
class(non_county_data) <- c('state', class(non_county_data))

```

### Extract State

For the county data, we want to extract the state. 

```{r }
#extract state names
county_data <- county_data %>%
    mutate(state = substr(area_name, nchar(area_name) -1, nchar(area_name)))

```

### Assign Divisions

For the state data, we want to assign divisions based on the divisions found [here](https://en.wikipedia.org/wiki/List_of_regions_of_the_United_States)

```{r }
## Assign divisions based on the state
#if there is no state assign "ERROR"

non_county_data <- non_county_data %>%
    mutate(division = case_when(area_name %in% toupper(c('Connecticut', 
                                                         'Maine', 
                                                         'Massachusetts', 
                                                         'New Hampshire', 
                                                         'Rhode Island', 
                                                         'Vermont')) ~ 'Division 1',
                                area_name %in% toupper(c('New Jersey', 
                                                         'New York',
                                                         'Pennsylvania')) ~ 'Division 2',
                                area_name %in% toupper(c('Illinois', 
                                                         'Indiana', 
                                                         'Michigan', 
                                                         'Ohio',
                                                         'Wisconsin')) ~ 'Division 3',
                                area_name %in% toupper(c('Iowa', 
                                                         'Kansas', 
                                                         'Minnesota', 
                                                         'Missouri', 
                                                         'Nebraska', 
                                                         'North Dakota', 
                                                         'South Dakota')) ~ 'Division 4',
                                area_name %in% toupper(c('Delaware', 
                                                         'Florida', 
                                                         'Georgia', 
                                                         'Maryland', 
                                                         'North Carolina', 
                                                         'South Carolina', 
                                                         'Virginia', 
                                                         'District of Columbia', 
                                                         'West Virginia')) ~ 'Division 5',
                                area_name %in% toupper(c('Alabama', 
                                                         'Kentucky', 
                                                         'Mississippi', 
                                                         'Tennessee')) ~ 'Division 6',
                                area_name %in% toupper(c('Arkansas', 
                                                         'Louisiana', 
                                                         'Oklahoma', 
                                                         'Texas')) ~ 'Division 7',
                                area_name %in% toupper(c('Arizona', 
                                                         'Colorado', 
                                                         'Idaho', 
                                                         'Montana', 
                                                         'Nevada', 
                                                         'New Mexico', 
                                                         'Utah', 
                                                         'Wyoming')) ~ 'Division 8',
                                area_name %in% toupper(c('Alaska', 
                                                         'California', 
                                                         'Hawaii', 
                                                         'Oregon', 
                                                         'Washington')) ~ 'Division 9',
                                TRUE ~ 'ERROR' )
                                )


```


## Data Processing with Functions

Now that we have gone through the process of transforming the data for a single data set, we want to create functions that will perform these actions on more than one data set with a similar structure.

### Select Columns and Pivot 

Again we will select the appropriate columns and pivot the data from wide to long, but this time creating a function

```{r}


select_and_pivot <- function(.data, var_name = 'Enrollment') {
    
#select area_name, STCOU, and columns ending with D
#pivot columns ending with D into 2 columns 
#one for the enrollment code and 
#one for the values
    out <- .data %>%
            select(area_name = Area_name, STCOU, ends_with('D')) %>%
            pivot_longer(cols = ends_with('D'), names_to = 'Code', values_to = var_name)
    
    return(out)
    
} 



```

### Extract Year and Measurement

Next we create a function to extract the measurement and year.

```{r}

extract_year_measurement <- function(.data, ...) {
    
#extract the year (digits 8 & 9) and format as a 4 digit year
#extract the measurement (the first 7 characters)  
    out <- .data %>%
         mutate(year = year(as.Date(substr(Code, 8,9), format = '%y')),
           measurement = substr(Code, 1, 7))
        
    return(out)
    
}


```

### Split Data

Then we create a function that will split the data into two data sets for county and state, and assign appropriate classes

```{r}
#if the area_name contains a comma assign it to a county data set
#otherwise assign it to a non_county data set
split_areas <- function(.data, ...) {
    
    #create a county and non-county column and 
    #then split the data into a list of 2 dataframes on that column
    out <- .data %>%
        mutate(name = if_else(str_detect(area_name, ','), 'County', 'Non_County')) %>%
        split(.$name) %>%
        map(select, -name)
        
    #assign classes to each dataset
    class(out$County) <- c('county', class(out$County))
    class(out$Non_County) <- c('state', class(out$Non_County))
   
   return(out)
    
}

```

### Extract State

We extract the state for the county data. Here we use the class of the data to help us identify the county data.

```{r}


extract_state <- function(.data, ...) {
    
    #assign the data to a variable called out
    out <- .data
    
    #if the class of the data is county then extract the state from the area
    if('county' %in% class(.data)) {
    
    out <- out %>%
    mutate(state = substr(area_name, nchar(area_name) -1, nchar(area_name)))
    
    } #otherwise do nothing
    
    return(out)

}


```


### Assign Divisions

Then assign divisions to the non-county data. Here we use the class of the data to identify the state data. 

```{r}

assign_division <- function(.data, ...) {
    
    #assign the data to a variable called out
    out <- .data
    
     #if the class of the data is stat then assign divisions based on the state
    if('state' %in% class(.data)) {
     
        out <- out %>%
        filter(!str_detect(area_name, ',' )) %>%
        mutate(division = case_when(area_name %in% toupper(c('Connecticut', 
                                                             'Maine', 
                                                             'Massachusetts', 
                                                             'New Hampshire', 
                                                             'Rhode Island', 
                                                             'Vermont')) ~ 'Division 1',
                                area_name %in% toupper(c('New Jersey', 
                                                         'New York',
                                                         'Pennsylvania')) ~ 'Division 2',
                                area_name %in% toupper(c('Illinois', 
                                                         'Indiana', 
                                                         'Michigan', 
                                                         'Ohio',
                                                         'Wisconsin')) ~ 'Division 3',
                                area_name %in% toupper(c('Iowa', 
                                                         'Kansas', 
                                                         'Minnesota', 
                                                         'Missouri', 
                                                         'Nebraska', 
                                                         'North Dakota', 
                                                         'South Dakota')) ~ 'Division 4',
                                area_name %in% toupper(c('Delaware', 
                                                         'Florida', 
                                                         'Georgia', 
                                                         'Maryland', 
                                                         'North Carolina', 
                                                         'South Carolina', 
                                                         'Virginia', 
                                                         'District of Columbia', 
                                                         'West Virginia')) ~ 'Division 5',
                                area_name %in% toupper(c('Alabama', 
                                                         'Kentucky', 
                                                         'Mississippi', 
                                                         'Tennessee')) ~ 'Division 6',
                                area_name %in% toupper(c('Arkansas', 
                                                         'Louisiana', 
                                                         'Oklahoma', 
                                                         'Texas')) ~ 'Division 7',
                                area_name %in% toupper(c('Arizona', 
                                                         'Colorado', 
                                                         'Idaho', 
                                                         'Montana', 
                                                         'Nevada', 
                                                         'New Mexico', 
                                                         'Utah', 
                                                         'Wyoming')) ~ 'Division 8',
                                area_name %in% toupper(c('Alaska', 
                                                         'California', 
                                                         'Hawaii', 
                                                         'Oregon', 
                                                         'Washington')) ~ 'Division 9',
                                TRUE ~ 'ERROR' )
                                )
    } #otherwise do nothing
    
     return(out)
     
}



```

### Combine Functions

Now we need a function that will perform the splitting of the data sets, extract the state, and assign divisions. We do this by using the functions created above in another function.

```{r}

#create a function that will use the above functions to
#1) split the data and assign classes
#2) extract the state for the county data 
#3) assign divisions for the non-county data
#note the map function in the purrr package is used to operate on a list of dataframes
#we have already used the classes of the datasets to identify which functions to apply to which datasets above
split_extract_assign <- function(.data, ...) {
    
    #split the data and assign classes
    splits <- split_areas(.data)
    
    #extract the state and assign divisions accordingly
   out <- splits %>%
       map(extract_state) %>%
       map(assign_division)
    
    return(out)
    
}

```


### Wrapper Function 

Now we put all of the steps together into one function call. Here we create a function that takes the url of the .csv file and performs all of the actions we have previously defined. 

```{r}

wrapper_function <- function(url,...) {
    
    #read in the data
    out_1 <- read.csv(url)
    #select columns and pivot from wide to long
    out_2 <- select_and_pivot(out_1,...)
    #extract year and measurement
    out_3 <- extract_year_measurement(out_2, ...)
    #split data into county and state, extract state, and assign divisions
    final_out <- split_extract_assign(out_3, ...) 
    
    #return the transformed data
    return(final_out)

    }
```


### Call Function

Now we will test out our function by calling it two data sets

```{r } 

EDU01a <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv")
    
EDU01b <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv")


``` 

### Combine Data

Lastly we combine the county data and the state data together by binding rows

```{r}


combine_data <- function(datasets) {
    
#bind rows of the county data
    County <- datasets %>%
        map(pluck, 'County') %>%
        bind_rows()
        
#bind rows of the state data      
    State <- datasets %>%
        map(pluck, 'Non_County') %>%
        bind_rows()
    
    #return a list containing the 2 combined data sets
    return(list('County' = County, 'State' = State))
    
    
}

#call the function on the data sets to combine into County and State
dat <- list(EDU01a, EDU01b) %>%
   combine_data()

```


## Create Generic Plot functions 

Because the base R plot function is created using the S3 object oriented system, different methods can be assigned to it that perform different actions based on the class of the object that is input into the function. We have already assigned the classes 'state' and 'county' to our data set objects. We will use these classes to create methods which tell the plot function how to operate on these classes. 

### Summarize State Data

First we want to create a plotting method for state data. We want to plot the mean value across each division and year. The data will need to be summarized by the year and division. We want a line plot with the year ont the x-axis, and the mean value (in this case mean of enrollment) on the y-axis.  

```{r}

plot.state <- function(.data, var_name = 'Enrollment') {
    
    out_data <- .data %>%
        filter(division != 'ERROR') %>%
        group_by(year, division) %>%
        summarise(mean = mean(get(var_name)))
        
        
    out_plot <- ggplot(out_data, aes(x = year, y = mean, color = division)) + 
        geom_line()
    
    return(out_plot)
    
        
}

#test the plot method on the state data
plot(dat$State, var_name = 'Enrollment')

```

### Summarize County Data

Now we create a plotting method for the county data. This time we want to allow the user to have more flexibility.

The user should be able to:  

* Specify the state
* Choose whether the top or bottom most counties will be shown
* Enter the number of top or bottom most values they are interested in

```{r}

plot.county <- function(.data, state_name = 'NC', var_name = 'Enrollment', order = 'top', num = 5) {
    
    #filter the data to the state entered by the user
    state_data <- .data %>%
        filter(state == state_name) 
    
    #get the means of each area for the state
    means <- state_data %>%
        group_by(area_name) %>%
        summarise(mean = mean(get(var_name))) 
    
    #if the user chooses the top most values arrange in descending order
    #limit to the number selected
    if(order == 'top') {
        
        plot_order <-  means %>%
            arrange(-mean) %>%
            head(num)
    }
    
    #if the user chooses the bottom most values arrange in ascending order
    #limit to the number selected
    if(order == 'bottom') {
        
        plot_order <- means %>%
            arrange(mean) %>%
            head(num)
        
    }
    
    #filter the data set to the top or bottom most areas accordingly
    plot_data <- state_data %>%
        filter(area_name %in% plot_order$area_name) 
    
    #plot the data
    out_plot <- ggplot(plot_data, aes(x = year, y = get(var_name), color = area_name)) + 
        geom_line() +
        ylab(var_name)
    
    return(out_plot)
    
}

#test the plot method on the county data
plot(dat$County, order = 'bottom', num = 7)

```

## Putting it All Together

Now that we have created our function and our generic plotting methods we want to test it out on our data with various conditions

### Testing the Functions and Methods

We will read in and transform the data sets again using our function

```{r}
#read in and prep data
EDU01a <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv")
    
EDU01b <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv")

```

Then combine the data into state and county

```{r}

#combine datasets into state and county
dat <- list(EDU01a, EDU01b) %>%
   combine_data()


```

And test various plotting parameters  

* Plot the state data set 

```{r}

plot(dat$State, var_name = 'Enrollment')

```

* Plot the county data set   
    
    - Specifying the state to be NC, the group to be top, and the number 10  
    - Specifying the state to be AZ, the group to be bottom, and the number 6  
    - Using default parameters  
    - Specifying the state to be OH, the group to be top, and the number 8  
    
    
```{r, fig.show="hold", out.width="50%"}

plot(dat$County, state_name = 'NC', order = 'top', num = 10)
plot(dat$County, state_name = 'AZ', order = 'bottom', num = 6)
plot(dat$County)
plot(dat$County, state_name = 'OH', order = 'top', num = 8)

```
    

### Read in more data

Now we will read in and transform four new data sets using our function.

```{r}

PST01a <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/PST01a.csv")
PST01b <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/PST01b.csv")
PST01c <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/PST01c.csv")
PST01d <- wrapper_function("https://www4.stat.ncsu.edu/~online/datasets/PST01d.csv")

```

Combine the data sets into County and State

```{r}
dat <- list(PST01a, PST01b, PST01c, PST01d) %>%
    combine_data()

```


And again test the plotting methods on various parameters: 

* Plot the state data set 

```{r}

plot(dat$State)

```

* Plot the county data set  

    - Specifying the state to be “PA”, the group to be top, and the number 5  
    – Specifying the state to be “TX”, the group to be bottom, and the number 12  
    – Using default parameters  
    – Specifying the state to be “NY”, the group being the top, and the number 6  

```{r , fig.show="hold", out.width="50%"}
plot(dat$County, state_name = 'PA', order = 'top', num = 5)
plot(dat$County, state_name = 'TX', order = 'bottom', num = 12)
plot(dat$County)
plot(dat$County, state_name = 'NY', order = 'top', num = 6)
```
    
    